---
title: "Kallisto bustools Processing"
output: html_notebook
---

This notebook is an example of processing a kallsto|bustools run as implemented in the `run-kallisto.nf` workflow.

```{r setup}
library(magrittr)
library(ggplot2)
library(SingleCellExperiment)
```


First we will download a data set to work with from S3. 
This set was quantified as part of a benchmarking set.
This step assumes that `awscli` is installed (which it is if you use the `scpca-r` docker image) and that you have already set up credentials.

```{r}
data_dir <- file.path("data", "kallisto", "834")
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)
quant_s3 <- "s3://nextflow-ccdl-results/scpca-benchmark/quants/kallisto/834-kallisto_cdna"
mito_s3 <- "s3://nextflow-ccdl-data/reference/homo_sapiens/ensembl-100/annotation/Homo_sapiens.ensembl.100.mitogenes.txt" 
mito_file <- file.path("data", "Homo_sapiens.ensembl.100.mitogenes.txt")
```

### Sync S3 files

```{r}
sync_call <- paste("aws s3 sync", quant_s3, data_dir, "--exclude 'bus/*'")
system(sync_call, ignore.stdout = TRUE)

# get the mitochondrial gene set
cp_call <- paste("aws s3 cp", mito_s3, mito_file)
system(cp_call)
```

## Read counts matrix
```{r}
file_base <- file.path(data_dir, "counts", "gene_count")

counts <- Matrix::readMM(paste0(file_base, ".mtx")) %>%
  # compress and transpose to genes in rows x cells in columns
  t() %>%
  as("dgCMatrix")
# add annotation info
dimnames(counts) <- list(readLines(paste0(file_base, ".genes.txt")),
                         readLines(paste0(file_base, ".barcodes.txt")))
```

## Identifying cells with droplets

### Cellranger default method first.

Based on library size, capturing cells with at least 10% of the UMI counts of the upper 99th percentile.

```{r}
system.time(drops <- DropletUtils::defaultDrops(counts))

sum(drops)
min(colSums(counts[,drops]))
```

### Knee plot method

Find the knee, keep the cells above that rank.

```{r}
system.time(rank_df <- DropletUtils::barcodeRanks(counts))
knee = metadata(rank_df)$knee
inflection = metadata(rank_df)$inflection

sum(rank_df$total >= knee)
min(colSums(counts[,which(rank_df$total >= knee)]))
```


```{r}
ggplot(as.data.frame(rank_df), aes(x = rank, y = total)) + 
  geom_point() +
  geom_hline(yintercept = knee, color = "blue") +
  geom_hline(yintercept = inflection, color = "darkgreen") +
  scale_x_log10() + 
  scale_y_log10()
```


### EmptyDrops method
Fancy method: difference from ambient RNA distribution.
(v. slow so may not be worth pursuing in this context, but could be added to a workflow?)
```{r}
system.time(drop_df <- DropletUtils::emptyDrops(counts))

is.cell <- drop_df$FDR <= 0.01 
sum(is.cell, na.rm=TRUE)
```

This method also keeps many more "cells", which may be net negative for downstream processing as well, but may also be worth pursuing for some kinds of analysis.

For now, we will proceed with the knee method to define cells.

## Make SCE object 

```{r}
cells <- which(rank_df$total >= knee)
sce <- SingleCellExperiment(list(counts = counts[ , cells]))
```

That results in a much more reasonably sized object, one we can work with as usual with SCE objects. 

## Simple QC
```{r}
mito_genes <- readLines(mito_file)

sce <- scater::addPerCellQC(
  sce, 
  subsets = list(mito = mito_genes[mito_genes %in% rownames(sce)])
) %>%
  scater::addPerFeatureQC()
```

```{r}
ggplot(as.data.frame(colData(sce)),
       aes(x = sum, y = detected, color = subsets_mito_percent)) + 
  geom_point(alpha = 0.1) + 
  scale_color_viridis_c()
```


### EmptyDrops QC

Out of curiousity, what do the EmptyDrops results look like?

```{r}
sce_ed <- SingleCellExperiment(list(counts = counts[ , which(is.cell)]))

sce_ed <- scater::addPerCellQC(
  sce_ed, 
  subsets = list(mito = mito_genes[mito_genes %in% rownames(sce_ed)])
) %>%
  scater::addPerFeatureQC()

ggplot(as.data.frame(colData(sce_ed)),
       aes(x = sum, y = detected, color = subsets_mito_percent)) + 
  geom_point(alpha = 0.1) + 
  scale_color_viridis_c()
```

What happens if we look at the knee plot after EmptyDrops excludes ambient expression drops?

```{r}
rank_df <- DropletUtils::barcodeRanks(counts(sce_ed))
knee = metadata(rank_df)$knee
inflection = metadata(rank_df)$inflection

sum(rank_df$total >= knee)
min(colSums(counts(sce_ed)[,which(rank_df$total >= knee)]))
```
Interestingly, it now recovers more cells, at a slightly lower threshold.


```{r}
ggplot(as.data.frame(rank_df), aes(x = rank, y = total)) + 
  geom_point() +
  geom_hline(yintercept = knee, color = "blue") +
  geom_hline(yintercept = inflection, color = "darkgreen") +
  scale_x_log10() + 
  scale_y_log10()
```


## Session Info
```{r}
sessionInfo()
```

