---
title: "Alevin-Fry Five Prime"
author: "Ally Hawkins for CCDL"
output: 
  html_notebook:
    toc: true
    toc_float: true
---

This notebook compares using Alevin-fry to pre process samples captured using the 5' kit from 10X rather than the usual 3' approach. 
I am comparing Alevin-fry, using the decided upon parameters we have chosen for analyzing the 3' libraries, to Cell Ranger for 2 5' samples. 

Alevin-fry is run using selective alignment to the splici index using the `cr-like-em` resolution for UMI deduplication. 
Additionally, to signify that we are aligning a 5' library, the library type argument was changed from `ISR` to `ISF`. 
The change in library type indicates how the reads are oriented and switches read 1 from the reverse strand to the forward strand. 
In contrast, Cell Ranger is able to automatically detect the library type based on how many of the reads are in the antisense direction. 

Here, we are testing how well Alevin-fry using the `-l ISF` option with 5' library performs in comparison to using Cell Ranger's automatic detection. 
To do this we will look at the distribution of UMI/cell, genes/cell, correlation of mean gene expression, and number of detected genes that overlap, similar to previous benchmarking comparisons. 

## Set Up
 

```{r}
library(magrittr)
library(ggplot2)
library(SingleCellExperiment)
library(ggupset)
```


```{r}
# path to results files with sces and qc dataframes from running benchrmarking_generate_qc_df.R
base_dir <- here::here()
file_dir <- file.path(base_dir, "data", "five-prime", "results")

# qc files 
quant_info_file <- file.path(file_dir, "quant_info.tsv")
coldata_df_file <- file.path(file_dir, "coldata_qc.tsv")
rowdata_df_file <- file.path(file_dir, "rowdata_qc.tsv")

# sce objects
fry_file <- file.path(file_dir, "splici_salign_cr-like-em_sces.rds")
cellranger_file <- file.path(file_dir, "cellranger_sces.rds")
```

```{r}
# read in dataframes needed for plotting
quant_info <- readr::read_tsv(quant_info_file)
coldata_df <- readr::read_tsv(coldata_df_file)
rowdata_df <- readr::read_tsv(rowdata_df_file)
```

```{r}
# read in sces
fry_sce <- readr::read_rds(fry_file)
cellranger_sce <- readr::read_rds(cellranger_file)
```

```{r}
alevin_fry_tools <- c("splici_salign_cr-like", "splici_salign_cr-like-em")

# merge coldata df with quant_info
coldata_info_df <- coldata_df %>%
  # rename tool to be either cellranger or alevin-fry
  dplyr::mutate(tool = ifelse(tool %in% alevin_fry_tools, "alevin-fry", "cellranger")) %>%
  dplyr::left_join(quant_info,
                   by = c("tool" = "tool", 
                          "quant_id" = "quant_dir"))
```

We are only interested in looking at cells that are found in both tools, so we restrict our analysis to cells that are present in both.  

```{r}
# filter for cells that are found in both alevin + cellranger
cell_counts <- coldata_info_df %>%  
  dplyr::count(cell_id, sample)

common_cells <- cell_counts %>%
  dplyr::filter(n == 2) %>%
  dplyr::pull(cell_id)

coldata_qc_common <- coldata_info_df %>%
  dplyr::filter(cell_id %in% common_cells) 
```

## Per Cell Comparisons

Here we are looking at the per cell metrics in just the shared cells across both alevin-fry and Cell Ranger.

```{r}
# mito content in shared cells
ggplot(coldata_qc_common, aes(x = tool, y = subsets_mito_percent, fill = tool)) + 
  geom_boxplot() + 
  facet_grid(~ sample) +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ylab("Mito Percent") + 
  xlab("")
```
```{r}
ggplot(coldata_qc_common, aes(x = tool, y = sum, fill = tool)) + 
  geom_boxplot() + 
  facet_grid(~ sample) +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ylab("UMI/cell") + 
  xlab("") + 
  ylim(c(0,40000))
```
```{r}
ggplot(coldata_qc_common, aes(x = tool, y = detected, fill = tool)) + 
  geom_boxplot() + 
  facet_grid(~ sample) +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ylab("Genes detected/cell") + 
  xlab("") + 
  ylim(c(0,7500))
```
Interestingly we see that alevin-fry shows higher mito content per cell and lower overall counts and number of genes detected per cell than Cell Ranger. 

### Shared Cells, Minimum Gene Coverage

We can also restrict the comparisons to only include genes that are found in at least 5% of cells to see if that low covered genes are affecting the distributions at all. 
To do that we will need to recalculate the per cell metrics using only genes found in > 5% of cells. 

```{r}
# combine all tools into one list
sce_list <- append(fry_sce, cellranger_sce)

# use addPerCellQC with gene detection threshold
# only genes with detection in > 5% of cells will be included in the QC calculations
add_qc_threshold <- function(sce, mito = mito_genes, threshold = 5){
  # remove previous colData first 
  colData(sce) <- NULL
  scpcaTools::add_cell_mito_qc(sce, mito = mito_genes, threshold = threshold)
}

# apply above function to all sces 
sce_threshold_list <- sce_list %>%
    purrr::map(add_qc_threshold)
```

```{r}
# merge back into a dataframe for plotting 
coldata_threshold <- sce_threshold_list %>%
  purrr::map_df(scpcaTools::coldata_to_df, .id = "quant_id") %>%
  # join with quant info to get sample and tool information 
  dplyr::left_join(quant_info, by = c("quant_id" = "quant_dir"))
```

```{r}
# look for shared cells only 
cell_counts_threshold <- coldata_threshold %>%  
  dplyr::count(cell_id, sample)

common_cells_threshold <- cell_counts_threshold %>%
  dplyr::filter(n == 2) %>%
  dplyr::pull(cell_id)

qc_common_threshold <- coldata_threshold %>%
  dplyr::filter(
    (cell_id %in% common_cells_threshold) 
  )
```

```{r}
ggplot(qc_common_threshold, aes(x = tool, y = sum, fill = tool)) + 
  geom_boxplot() + 
  facet_grid(~ sample) +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ylab("UMI/cell") + 
  xlab("") + 
  ylim(c(0,40000))
```
```{r}
ggplot(qc_common_threshold, aes(x = tool, y = detected, fill = tool)) + 
  geom_boxplot() + 
  facet_grid(~ sample) +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ylab("Genes detected/cell") + 
  xlab("") + 
  ylim(c(0,1000))
```
It appears that restricting to highly covered genes does not improve the distributions of UMI/cell and genes detected/cell and we see they are still much lower than with Cell Ranger. 

## Per Gene Comparison

Next, we will compare the mean gene expression of shared genes. 
Here we calculate the correlation coefficient of mean gene expression and we can plot the relationship of mean gene expression between the two tools. 

```{r}
# combine rowdata with quant info
rowdata_info_df <- rowdata_df %>%
  dplyr::mutate(tool = dplyr::case_when(tool == "splici_salign_cr-like-em" ~ "alevin-fry",
                                        tool == "cellranger" ~ "cellranger")) %>%
  dplyr::left_join(quant_info,
                   by = c("tool" = "tool", 
                          "quant_id" = "quant_dir"))


gene_counts <- rowdata_info_df %>% 
  # remove genes that have a low frequency of being detected
  dplyr::filter(detected >= 5.0) %>%
  dplyr::count(gene_id, sample)

# restrict to only common genes 
common_genes <- gene_counts %>%
  dplyr::filter(n == 2) %>%
  dplyr::pull(gene_id)

rowdata_qc_common <- rowdata_info_df %>%
  dplyr::filter(
    (gene_id %in% common_genes) 
  )
```

```{r}
# spread table for comparisons
rowdata_cor <- rowdata_qc_common %>%
  dplyr::select(tool, gene_id, sample, mean) %>%
  # spread the mean expression stats to one column per caller
  tidyr::pivot_wider(id_cols = c(gene_id, sample),
                     names_from = c("tool"),
                     values_from = mean) %>%
  # drop rows with NA values to ease correlation calculation
  tidyr::drop_na()
```


```{r}
# calculate spearman correlation for alevin fry vs. cellranger for each sample
rowdata_cor %>% 
  dplyr::group_by(sample) %>%
  dplyr::summarize(
    alevin_fry_cellranger_cor = cor(`cellranger`, `alevin-fry`, method = "spearman")
  )
```
```{r}
ggplot(rowdata_cor, aes(x = `cellranger`, y = `alevin-fry`)) +
  geom_point(size = 0.5, alpha = 0.1) + 
  facet_wrap(~ sample) + 
  scale_x_log10() + 
  scale_y_log10() + 
  labs(x = "Cell Ranger mean gene expression", y = "Alevin Fry Mean gene expression") + 
  theme_classic()
```
Unlike with the 3' libraries, the 5' libraries have fairly poor correlation between Alevin-fry and Cell Ranger. 
It looks like there is more of a blob shape rather than correlation along the diagonal, and for SCPCR000287, Cell Ranger tends to give a higher gene expression for most genes than with Alevin-fry. 


## Overlapping Genes 

In addition to looking at mean gene expression, we can also look at the actual genes that are detected in Cell Ranger and Alevin-fry and identify how much of the gene sets are overlapping or not. 
Because they are using different ensembl assemblies for their indices, we will need to first restrict the analysis to only genes that are able to be detected in both tools. 


```{r}
# make a function to filter sces by subset of cells and re-calculate feature stats
filter_sce <- function(sce, cells){
  cells_to_keep <- colnames(sce) %in% cells
  SummarizedExperiment::rowData(sce) <- NULL
  sce[, cells_to_keep] %>%
    scater::addPerFeatureQC()
}
```

```{r}
# filter all sces together 
all_sces <- list(fry_sce,
                 cellranger_sce)
names(all_sces) <- c('alevin-fry', 'cellranger')

sces_filtered <- all_sces %>%
  purrr::map(
    ~ purrr::map(.x, filter_sce, cells = common_cells)
  )

# grab rowdata from filtered sces 
rowdata_filtered <- sces_filtered %>% 
  purrr::map_df(
    ~ purrr::map_df(.x, scpcaTools::rowdata_to_df, .id = "quant_dir"),
    .id = "tool"
  ) %>%
  dplyr::left_join(quant_info)
```
```{r}
# get genes common in all tools
common_genes <- rowdata_filtered %>%
  dplyr::select(gene_id, tool) %>%
  dplyr::distinct() %>%
  dplyr::group_by(gene_id) %>%
  dplyr::tally() %>%
  # filter for genes found in both cellranger and alevin-fry
  dplyr::filter(n == 2) %>%
  dplyr::pull(gene_id)

# filter rowdata_df to only include genes found in all tools 
rowdata_filtered <- rowdata_filtered %>% 
  dplyr::filter(gene_id %in% common_genes)
```


Now that we have restricted to only common genes, we can then compare genes that were detected in one tool but not the other. 
To do this, we only want to include genes that have a mean > 0 and are detected in more than 0 cells. 

```{r}
# filter
rowdata_filtered <- rowdata_filtered %>%
  dplyr::filter(mean > 0 & detected > 0)
```

```{r}
gene_detect_df <- rowdata_filtered %>%
  dplyr::group_by(gene_id) %>%
  dplyr::summarise(tools_detected = list(unique(tool)))

ggplot(gene_detect_df, aes(x = tools_detected)) +
  geom_bar() +
  scale_x_upset(n_intersections = 3)
```
It appears that despite the low correlation in gene expression, and low similarity in UMI counts/ cell, the number of overlapping detected genes between Alevin-fry and Cell Ranger is quite high. 
The majority of genes are being identified in both tools, although the previous analysis suggests they are being identified at different levels. 

## Session Info 

```{r}
sessioninfo::session_info()
```

