---
title: "Genetic demultiplexing vs. cellhash"
author: "Josh Shapiro for CCDL"
output: 
  html_notebook:
    toc: true
    toc_float: true
params:
  lib_id: "SCPCL000535"
---

# Set Up
 
Libraries:

```{r setup}
suppressPackageStartupMessages({
  library(ggplot2)
  library(SingleCellExperiment)
})

set.seed(2021)
```

File paths:

```{r}
file_dir <- file.path("data", "multiplex")


# singleCellExperiment files from 
filtered_file <- file.path(file_dir, paste0(params$lib_id, "_filtered.rds"))


## outfiles
outdir <- "results"
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)
demux_tsv <- file.path(outdir, paste0(params$lib_id, "_demux.tsv"))
demux_tables <- file.path(outdir, paste0(params$lib_id, "_demux_table.txt"))

plotdir <- "plots"
demux_umap_plot <- file.path(plotdir, paste0(params$lib_id, "_demux_umap.png"))
```

## Functions

```{r}
# calculate the agreement between two lists, with and or without NAs included
calc_agreement <- function(call_a, call_b, na.include = FALSE){
  agree <- sum(call_a == call_b, na.rm = TRUE)
  disagree <- sum(call_a != call_b, na.rm = TRUE)
  if(na.include){
    # add single NA values
    disagree <- disagree  + sum(xor(is.na(call_a), is.na(call_b)))
  } 
  return(agree/(agree + disagree))
}

```




# Single cell processing and filtering

The SingleCellExperiment files have already been processed via the `scpca-nf` pipeline, which performs filtering and adds cellhash-based demultiplexing calls using `DropletUtils::HashedDrops` and `Seurat::HTODemux`.

The pipeline also incorporates genetic information inferred from bulk sequencing and performs genetic demultiplexing with [cellsnp](https://cellsnp-lite.readthedocs.io/en/latest/) and [Vireo](https://vireosnp.readthedocs.io/en/latest/).


```{r}
sce <- readRDS(filtered_file)
```

## Sample information

The metadata in the processed SingleCellExperiment includes information about the samples included in this library:

```{r}
metadata(sce)$sample_id
```


# Demultiplexing results

The confident calls (by default measures) are stored in the `colData` data frame, prefixed by method, so we will extract those calls for comparisons.
Doublets here are coded as part of the `NA` group.

```{r}
demux_calls <- data.frame(
  hashedDrops = sce$hashedDrops_sampleid,
  HTODemux = sce$HTODemux_sampleid,
  vireo = sce$vireo_sampleid
)

# Summarize all of the calls
demux_calls |>
  dplyr::mutate_all(as.factor) |>
  summary()
```

In general, we see that `DropletUtils::hashedDrops` tends to make many fewer calls than `Seurat::HTODemux`. 
In either case, we have a large excess of `NA` calls, where the cellhash-based demultiplexing algorithms did not find a single best call. 

This may partly be a function of the quality of the cellhash results for this particular library.
As seen in `14-cellhash-exploration.Rmd`, the cellhash data for these samples seems to have a lot of background signal that may make accurate demultiplexing based on cellhash data alone difficult.

The genetic demultiplexing calls made by `vireo` seem to identify the most cells, but the relative frequencies of the samples do not necessarily agree with the results from `HTODemux`.
This could be because different sets of cells are being called by the two methods, or because the actual calls differ, so we can look in a bit more detail.

## Comparing classifications

To compare the calls between methods, we will build some contingency tables among the call sets. 


```{r}
ds_table <- table(demux_calls$hashedDrops, 
                  demux_calls$HTODemux, 
                  useNA = "always", 
                  dnn = c("hashedDrops", "HTODemux"))
ds_table
```


```{r}
ds_agreement <- calc_agreement(demux_calls$hashedDrops, demux_calls$HTODemux)
ds_agreement_withNA <- calc_agreement(demux_calls$hashedDrops, demux_calls$HTODemux, na.include = TRUE)

glue::glue("
  hashedDrops vs. HTODemux
  Agreement: {sprintf(ds_agreement, fmt ='%.3f')}
  With non-calls: {sprintf(ds_agreement_withNA, fmt ='%.3f')}
")
```


```{r}
dv_table <- table(demux_calls$hashedDrops, 
                  demux_calls$vireo, 
                  useNA = "always", 
                  dnn = c("hashedDrops", "vireo"))
dv_table
```
```{r}
dv_agreement <- calc_agreement(demux_calls$hashedDrops, demux_calls$vireo)
dv_agreement_withNA <- calc_agreement(demux_calls$hashedDrops, demux_calls$vireo, na.include = TRUE)

glue::glue("
  hashedDrops vs. vireo
  Agreement: {sprintf(dv_agreement, fmt ='%.3f')}
  With non-calls: {sprintf(dv_agreement_withNA, fmt ='%.3f')}
")
```


```{r}
sv_table <- table(demux_calls$HTODemux, 
                  demux_calls$vireo, 
                  useNA = "always", 
                  dnn = c("HTODemux", "vireo"))
sv_table
```
```{r}
sv_agreement <- calc_agreement(demux_calls$HTODemux, demux_calls$vireo)
sv_agreement_withNA <- calc_agreement(demux_calls$HTODemux, demux_calls$vireo, na.include = TRUE)

glue::glue("
  HTODemux vs. vireo
  Agreement: {sprintf(sv_agreement, fmt ='%.3f')}
  With non-calls: {sprintf(sv_agreement_withNA, fmt ='%.3f')}
")
```

Overall, when there are calls to be made, agreement among methods is quite good.
However, the proportion of non-calls is extremely high, and can be quite non-overlapping.

Without any way to determine ground truth, it is extremely difficult to assess which of the methods is really doing best in this case.


## Visualizing assignments

Before visualization, we will do some more  filtering so as to remove cells with little expression information and those with high mitochondrial content. 

```{r}
sce <- sce[, which(sce$detected > 500 & sce$subsets_mito_percent < 10)]
```

Now we perform some quick normalization, and generate a UMAP (with default parameters, as we are just having a quick look).


```{r}
qclust <- scran::quickCluster(sce)
sce <- sce |> 
  scran::computeSumFactors(clusters = qclust) |>
  scater::logNormCounts() |>
  scater::runUMAP()
```

Rearrange the data for easier plotting: 
```{r}
umap_data <- as.data.frame(reducedDim(sce, "UMAP"))
colnames(umap_data) <- c("UMAP1", "UMAP2")

# create entries for each method's calls for each cell & UMAP coordinates
umap_plot_data <- dplyr::bind_rows(
  hashedDrops = dplyr::mutate(umap_data, sample = sce$hashedDrops_sampleid),
  HTODemux = dplyr::mutate(umap_data, sample = sce$HTODemux_sampleid),
  vireo = dplyr::mutate(umap_data, sample = sce$vireo_sampleid),
  .id = "method"
) |>
  dplyr::arrange(!is.na(sample)) # move NA to top for plotting in back
```

Make a plot of all assignments
```{r}
ggplot(umap_plot_data, aes(x = UMAP1, y = UMAP2, color = sample)) + 
  geom_point(size = 1, alpha = 0.2) + 
  scale_color_brewer(palette = "Dark2", na.value = "grey70") +
  guides(color = guide_legend(override.aes = list(size = 2,
                                                  alpha = 1))) + 
  facet_wrap("method", nrow = 2) +
  theme_bw() + 
  theme(aspect.ratio = 1,
        axis.text = element_blank(), 
        axis.ticks = element_blank())

```
Overall, it does look like the `vireo` results are most consistent with the overal structure of the data, and thus seem most likely to reflect the true sample identities.
Of course, we do not know the true identities, and it is possible that the genetic demultiplexing is somehow conflating genetic and expression variation, but that seems less likely. 


## Doublet calls

The demultiplexing methods can potentially identify doublets, where more than one cell of different types ended up in the same droplet. 
Here we will look at the overall rate of doublet calls for each method.

```{r}
# select doublet-related information for each method
demux_calls_doublet <- data.frame(
  hashedDrops_doublet = altExp(sce, "cellhash")$hashedDrops_Doublet,
  hashedDrops_conf = altExp(sce, "cellhash")$hashedDrops_Confident,
  HTODemux_class = altExp(sce, "cellhash")$HTODemux_classification.global,
  vireo_calls = sce$vireo_donor_id
)
demux_classes <- demux_calls_doublet |> 
  dplyr::mutate( # make the _class calls match HTODemux classes
    hashedDrops_class = dplyr::case_when(
      hashedDrops_doublet ~ "Doublet",
      hashedDrops_conf ~ "Singlet",
      TRUE ~ "Negative"
    ),
    vireo_class = dplyr::case_when(
      vireo_calls == "doublet" ~ "Doublet",
      vireo_calls == "unassigned" ~ "Negative",
      is.na(vireo_calls) ~ "Negative",
      TRUE ~ "Singlet"
    )
  ) |>
  dplyr::select("hashedDrops_class", "HTODemux_class", "vireo_class") |>
  dplyr::mutate_all(as.factor)

summary(demux_classes)
```

Here we see that `HTODemux` seems to call the most doublets, which again makes sense if the cellhash background signal is substantial. 


# Conclusions

The main challenge we see in this data is that there seems to be very high background in the cellhash/HTO signal in these data.
The high background levels make it hard to pick out the sample-specific signal, so the HTO-based methods do not seem to perform very well.

`DropletUtils::hashedDrops` seems to be the most conservative method, and identifies very few cells overall. 

`Seurat::HTODemux()` is more aggressive, calling ~100x more cells, but those calls may not be particularly reliable in this case. 
Looking at the placement on the UMAP plots, they seem to span across clusters fairly broadly, which could be fine if we did not expect there to be any clustering by sample, but it is not clear what the expectation should be. 

`vireo`, which bases its calls on the genetic variants detected in each cell shows a stronger correspondence between sample calls and the expression variation, which makes sense if each sample is likely to have cells with distinct expression profiles.
There are still clusters with cells from multiple samples, which also seems like it should be expected. 

Overall, it remains unclear which if any of the methods is the most reliable, and attempting to answer that question  is beyond the scope of this analysis, and indeed may be beyond the scope of this project as a whole.
Rather than presenting a single set of calls as definitive and providing demultiplexed libraries for download, it seems like the best course of action is to allow downloading of the multiplexed library with all samples included, along with the calls and statistics generated by each of the demultiplexing methods.
The task of deciding among the methods will be left to the end user, using the included statistics to allow for adjusting any thresholds and recalling cells as they deem fit.
For the HTO-based methods, all of the count data will also be present to allow the application other demultiplexing methods.


# Session info
```{r}
sessioninfo::session_info()
```
