---
title: "Compare eisaR spliced & ensembl transcriptomes"
output: html_notebook
---

## Setup

### Load Libraries for import

```{r setup}
library(Biostrings)
library(BSgenome)
library(eisaR)
library(GenomicFeatures)
library(tidyverse)
library(rtracklayer)
```

### File and Directory Setup

```{r}
## get file directories
ref_dir <- file.path("..", "..", "workflows", "rnaseq-ref-index")

```

### Load gtf files for comparisons

```{r}
# load in the gtf files
spliced_gtf_file <- file.path(ref_dir, "annotation", "Homo_sapiens.GRCh38.103.spliced.txome.gtf.gz")
ensembl_gtf_file <- file.path(ref_dir, "annotation", "Homo_sapiens.GRCh38.103.gtf.gz")

spliced_gtf <- rtracklayer::import(spliced_gtf_file) %>% as.data.frame()
ensembl_gtf <- rtracklayer::import(ensembl_gtf_file) %>% as.data.frame()
```

## GTF Comparison

Start by getting just the transcript ids for each set, and compare them.

```{r}
spliced_gtf_txs <- spliced_gtf %>%
  drop_na(transcript_id) %>%
  pull(transcript_id) %>%
  unique() %>% 
  sort()

ensembl_gtf_txs <- ensembl_gtf %>%
  drop_na(transcript_id) %>%
  pull(transcript_id) %>%
  unique() %>% 
  sort()

# Are these the same? 
all.equal(spliced_gtf_txs, ensembl_gtf_txs)
```
So every transcript in one GTF seems to appear in the other!

## Fasta comparison
Lets look at the fasta files too...

```{r}
spliced_fasta <- file.path(ref_dir, "fasta", "Homo_sapiens.GRCh38.103.spliced.txome.fa.gz")
ensembl_fasta <- file.path(ref_dir, "fasta", "Homo_sapiens.GRCh38.txome.fa.gz")

spliced_txome <- Biostrings::readDNAStringSet(spliced_fasta)
ensembl_txome <- Biostrings::readDNAStringSet(ensembl_fasta)
```

How long is each?
```{r}
glue::glue("
  The eisaR transcriptome has:   {length(spliced_txome)} transcripts.
  The ensembl transcriptome has: {length(ensembl_txome)} transcripts.
  ")
```

So there are more in the ensembl fasta file, but what are they?
```{r}
spliced_fa_txs <- stringr::word(names(spliced_txome), 1)
ensembl_fa_txs <- stringr::word(names(ensembl_txome), 1) %>%
  stringr::word(1, sep = fixed('.')) # strip version
```

The vast majority are shared.
```{r}
shared_fa_tx <- intersect(spliced_fa_txs, ensembl_fa_txs)
length(shared_fa_tx)
```

Some are apparently unique to the spliced fasta file.
```{r}
spliced_only <- ! spliced_fa_txs %in% shared_fa_tx
sum(spliced_only)

head(spliced_txome[spliced_only])
```
All of these are in the spliced gtf:
```{r}
sum(names(spliced_txome)[spliced_only] %in% spliced_gtf_txs)
```

What are they?

```{r}
spliced_gtf %>%
  filter(transcript_id %in% names(spliced_txome)[spliced_only],
         type == 'transcript')
```


### Ensembl only

Some are unique to the ensembl fasta.
```{r}
ensembl_only <- ! ensembl_fa_txs %in% shared_fa_tx 

sum(ensembl_only)
head(ensembl_txome[ensembl_only])
```

These seem to be missing from the gtf file!
```{r}
sum(ensembl_fa_txs[ensembl_only] %in% ensembl_gtf_txs)
```

Since the Ensembl fasta file encodes more info in the names, we can look at that:

```{r}
info_fields = c('transcript', 
                'type', 
                'location', 
                'gene',
                'gene_biotype', 
                'transcript_biotype', 
                'gene_symbol',
                'description')
missing_tx_info <- data.frame(transcripts = names(ensembl_txome[ensembl_only])) %>%
  separate(transcripts, 
           sep = ' ',
           into = info_fields,
           extra = 'merge') %>%
  mutate(location = str_replace(location, "chromosome:GRCh38:", ""),
         gene = str_replace(gene, "gene:", ""),
         gene_biotype = str_replace(gene_biotype, "gene_biotype:", ""),
         transcript_biotype = str_replace(transcript_biotype, "transcript_biotype:", ""),
         gene_symbol = str_replace(gene_symbol, "gene_symbol:", ""),
         description = str_replace(description, "description:", ""),
         )
```

At a glance, it looks like all of these are on "extra" chromosomes. Is that true?

```{r}
missing_tx_info %>%
  mutate(chromosome = word(location, 1, sep = ':')) %>%
  count(chromosome) %>%
  arrange(desc(n))
```
It certainly seems to be.

Should we add the non-primary assembly to our references, or are we okay with these differences?


## Splici vs. Spliced_Intron Fasta
In generating the splici index, there is a step to obtain unique sequences only before writing the sequences to the fasta. 
This results in the size of the splici.fa being much smaller than the spliced_intron.fa, suggesting that they are different and that the spliced_intron.fa contains duplicated sequences.

```{r}
# look at splici vs. spliced_intron fasta
splici_fasta <- file.path(ref_dir, "fasta", "Homo_sapiens.GRCh38.103.splici.fa.gz")
spliced_intron_fasta <- file.path(ref_dir, "fasta", "Homo_sapiens.GRCh38.103.spliced_intron.txome.fa.gz")

splici_txome <- Biostrings::readDNAStringSet(splici_fasta)
spliced_intron_txome <- Biostrings::readDNAStringSet(spliced_intron_fasta)
```

```{r}
# get transcript ID's
splici_fa_txs <- stringr::word(names(splici_txome),1)
spliced_intron_fa_txs <- stringr::word(names(spliced_intron_txome),1)
```

```{r}
# how many of the spliced_intron transcript ID's are unique? 
unique_spliced_intron <- unique(spliced_intron_fa_txs)

# what's the intersect between the spliced_intron and splici transcript ID's? 
shared_splici_fa_tx <- intersect(splici_fa_txs, spliced_intron_fa_txs)

glue::glue("
  The splici transcriptome has: {length(unique_spliced_intron)} unique transcript IDs.
  The splici transcriptome and spliced_intron transcriptome have: {length(shared_splici_fa_tx)} shared transcript IDs.
  ")
```


This tells me that the spliced_intron.fasta contains duplicated sequences that map to multiple transcript ID's. 
How 

## Session info

```{r paged.print=FALSE}
sessioninfo::session_info()
```
